\textbf{The Illusion Game: A Novel Experimental Paradigm to study Visual Illusions}

### 2022-09-02 manuscript.Rmd

Visual illusions are fascinating stimuli capturing a key feature of our neurocognitive systems. They eloquently show that our brains do not simply map veridical accounts of physical reality; we generate perceptions of the environment by taking into consideration the contextual information and the prior knowledge that manifests in our subjective conscious experience [@carbon2014]. Despite the withstanding interest in illusions within various fields of study, including that of visual perception [@day1972; @eagleman2001; @gomez-villa2022 ], consciousness science [@caporuscio2022, @lamme2020] and psychiatry [@notredame2014; @gori2016; @razeghi2022; @teufel2015re], several gaps still remain in the present research. Most notably, there is currently no universally agreed upon neurocognitive mechanism that explains individuals' susceptibility to visual illusions [@mylniec2016]. For instance, while some researchers have tried to explain our sensitivity to illusory effects as a result of deficits in the low-level visual processing system [@cretenoud2019; @gori2016], others have provided a compelling case using a top-down approach, suggesting that such visual phenomena occur as a result of a conflict between our visual input and our prior beliefs[@teufel2018; @caporuscio2022]. Furthermore, results from studies that have been conducted to elucidate the mechanism underlying our susceptibility towards visual illusions remain relatively mixed. Whereas higher resistance towards such illusions have been reported for individuals with pathologically strong prior beliefs (such as schizophrenics) and atypical sensory perception (for example, those with autism spectrum disorder (ASD)) [@notredame2014; @park2022; @giaouri2011; @keane2014], other studies have found no significant differences between such individuals and healthy controls [@spencer2014; @yang2012; @tibber2013; @kaliuzhna2019] or only a weak correlation between the magnitude of visual illusions and such individuals' susceptibility to these illusory effects [@grzeczkowski2018; @manning2017].

@@

<!-- Target, strength, difficulty -->
Indeed, many visual illusions can be seen as being composed of *targets* (e.g., same-length lines), of which perception is biased by the *context* (e.g., in the Müller-Lyer illusion, the same-length line segments appear to have different lengths when they end with inwards or outwards pointing arrows). In past studies, the implemented paradigms have mostly involved prompting the subjective extent the context biases the perception (for instance, by getting participants to adjust the parameters of one comparison stimulus (the size of a circle) to another (circle) e.g., @mylniec2016a; @grzeczkowski2018). Alternatively, *Pyllusion* allows us to create illusions while manipulating specific features of interest (e.g., size, length etc., ) and thereby the difficulty in distinguishing them, so that the targets are objectively distinct (i.e., one line segment is physically longer than the other), and the strength of the generated illusion are variable (i.e., the biasing angle of the arrows are more or less acute).

<!-- Paradigm -->
This opens the door of an experimental task in which participants have to make perceptual judgments about the targets (e.g., which segment in the longest) under different conditions of objective difficulty and illusion strength. Moreover, the illusion effect can be either "incongruent" (making the task even harder by biasing the perception in the opposite way) or "congruent" (making the task easier). Although visual illusions are inherently tied to subjective perception, this framework allows a reversal of the traditional paradigm that could quantify the "objective" effect of illusions by measuring its behavioral effect (error rate and reaction times) on the performance in a perceptual task.

<!-- Goals -->
In the present set of preregistered studies, we will first test this novel paradigm by investigating if the effect of illusion and task difficulty can be manipulated continuously, and separated statistically. Then, we will further utilize the paradigm to assess whether 10 different cannonical illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) share a common latent factor (a long-standing debate, [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]). Finally, we will investigate how the the inter-individual sensitivity to illusions relates to other characteristics, such as demographic variables and personality.

@@

The objective of Study 1 is to act as a pilot study in order to gather some preliminary data to 1) assess whether the stimuli generated by *Pyllusion* behaves as expected for each of the 10 illusion types (i.e., increase of task difficulty and illusion strength leads to an increased error rate) and 2) develop a gauged understanding of the magnitude of effects, to then refine the stimuli parameters to a more sensible range (i.e., not overly easy and not impossibly hard).

@@

Fifty-two participants were recruited via *Prolific* ([www.prolificacademic.co.uk](www.prolificacademic.co.uk)), a popular crowd-sourcing platform to recruit subjects for online behavioural research. In a study comparing across different research recruitment platforms (e.g., MTurks, Qualtrics etc.,,), participants from *Prolific* were found to provide the highest data quality across all examined measures, including that of attention (the extent to which participants actually read the questions prior to responding), comprehension (the extent to which participants were able to understand the task instructions or items), honesty (the degree to which participants responded to questions truthfully) and reliability (the extent to which participants' responses are internally consistent within the same scale) [@peer2022]. For the purposes of our study, no restrictions on age or country were imposed on participant recruitment; the only requisite was a "fluent" proficiency in English to ensure that the task instructions, given in English, would be well-comprehended. Participants were incentivized with a reward of about \textsterling 7.5 for completing the task, which took an average of an hour to finish.

@@

Using the same recruitment procedure as in study 1, we recruited 256 participants, within which 6 were identified as outliers and excluded, leaving a final sample of 250 participants (Mean age = 26.5, SD = 7.6, range: [18, 69]; Sex: 48% females, 52% males). Please see the next study (study 3) for the full demographic breakdown. We discarded blocks with more than 50% of errors (2.16% of trials) and 0.76% trials with extreme response times (\< 125 ms or \> 4 SD above mean).

@@

The first part of the analysis focuses on modelling the effect of illusion strength and task difficulty on errors and reaction time, within each illusion. In order to achieve that, we will start by fitting a General Additive Models (GAMs), which can accommodate possible non-linear effects and interactions. Errors were analyzed using Bayesian binomial mixed models, and reaction times (RT) of correct responses were analyzed using an ex-Gaussian family with the same fixed effects entered for the location $\mu$ (mean), scale $\sigma$ (spread) and tail-dominance $\tau$ of the RT distribution [@balota2011moving; @matzke2009psychological].

@@

<!-- This is likely to be caused by the outline. It will be fixed in the next release of the Pyllusion software. -->
Future studies should investigate reaction times to find the most appropriate models, and / or use models that integrate errors and reaction time (e.g., drift diffusion models).

@@

This study was based on the data collected in study 2. However, the variables of interest here were taken from the questionnaires that were inserted in between the two series of illusion blocks. We used the *IPIP6* (24 items; [@sibley2011]) to measure 6 personality traits (namely, extraversion, openness, conscientiousness, agreeableness, neuroticism and honesty-humility). On the other hand, the PID-5 (25 items; [@hopwood2012]) was used to measure the degree to which participants inhibit psychopathological personality traits (Disinhibition, antagonism, detachment, negative affect and psychoticism). The participants were the same as in study 2. However, due to a technical issue, no data on the personality questions are available for the first eight participants.

@@

Using the parametric illusion generation framework we developed, *Pyllusion* [@makowski2021a], we have hence shown that illusions can be manipulated continuously across several different visual illusions. This thus opens the door for new illusions-based paradigms and tasks, therefore making it possible for future researchers to directly manipulate specific features and parameters of the illusion that are of interest. The validation of this novel framework also affords future illusion scientists a standardized measure of illusion susceptibility, instead of relying on conventional methods that depend upon participants' subjective perceptions. In our paradigm, in which we apply this approach to a reaction-time task, we were able to measure inter-individual scores of objective illusion sensitivity.

@@

We strongly invite researchers to explore and re-analyze our dataset with other approaches and methods to push the understanding of visual illusions and illusion sensitivity further. The task, data and analysis script are available in open-access at **https://github.com/RealityBending/IllusionGameValidation**.

>>>>

Visual illusions are fascinating stimuli capturing a key feature of our neurocognitive systems. They elegantly demonstrate that our brains do not simply map veridical accounts of physical reality; we generate perceptions of the environment by taking the contextual information into consideration, and the prior knowledge that manifests in our subjective conscious experience [@carbon2014]. Despite the withstanding interest in illusions within various fields of study, including that of visual perception [@day1972; @eagleman2001; @gomez-villa2022 ], consciousness science [@caporuscio2022, @lamme2020] and psychiatry [@notredame2014; @gori2016; @razeghi2022; @teufel2015re], several gaps still remain in the present research. Most notably, there is currently no universally agreed upon neurocognitive mechanism that explains individuals' susceptibility to visual illusions [@mylniec2016]. For instance, while some researchers have tried to explain our sensitivity to illusory effects as a result of deficits in the low-level visual processing system [@cretenoud2019; @gori2016], others have provided a compelling case using a top-down approach, suggesting that such visual phenomena occur as a result of a conflict between our visual input and our prior beliefs[@teufel2018; @caporuscio2022]. Furthermore, results from studies that have been conducted to elucidate the mechanism underlying our susceptibility towards visual illusions remain relatively mixed. Whereas higher resistance towards such illusions have been reported for individuals with pathologically strong prior beliefs (such as schizophrenics) and atypical sensory perception (for example, those with autism spectrum disorder [ASD]) [@notredame2014; @park2022; @giaouri2011; @keane2014], other studies have found no significant differences between such individuals and healthy controls [@spencer2014; @yang2012; @tibber2013; @kaliuzhna2019] or only a weak correlation between the magnitude of visual illusions and such individuals' susceptibility to these illusory effects [@grzeczkowski2018; @manning2017].

@@

<!-- Target, strength, difficulty -->

Indeed, many visual illusions can be seen as being composed of *targets* (e.g., same-length lines), of which perception is biased by the *context* (e.g., in the Müller-Lyer illusion, the same-length line segments appear to have different lengths when they end with inwards or outwards pointing arrows). Past studies have implemented paradigms emphasizing the subjective perception of the participant, to examine the extent to which individual context biases perceptual variance (for instance, having participants adjust the parametres of the target stimulus to a reference stimulus relying only on their perception, eg., @mylniec2016a; @grzeczkowski2018). Alternatively, Pyllusion allows us to create illusions while manipulating specific features of interest (e.g., size, length etc.), and thereby the difficulty in distinguishing them, so that the targets are objectively distinct (i.e., one line segment is physically longer than the other), and the strength of the generated illusion are variable (i.e., the biasing angle of the arrows are more or less acute).

<!-- Paradigm -->

This opens the door of an experimental task in which participants have to make perceptual judgments about the targets (e.g., which segment in the longest) under different conditions of objective difficulty and illusion strength. Moreover, the illusion effect can be either "incongruent" (making the task even harder by biasing the perception in the opposite way) or "congruent" (making the task easier). Although visual illusions are inherently tied to subjective perception, this framework allows a reversal of the traditional paradigm that could quantify the "objective" effect of illusions by measuring its behavioral effect (error rate and reaction times) on the performance in a perceptual task.

<!-- Goals -->

In the present set of preregistered studies, we will first test this novel paradigm by investigating if the effect of illusion and task difficulty can be manipulated continuously, and separated statistically. Then, we will further utilize the paradigm to assess whether 10 different canonical illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) share a common latent factor (a long-standing debate, [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]). Finally, we will investigate how the the inter-individual sensitivity to illusions relates to other characteristics, such as demographic variables and personality.

@@

The objective of Study 1 is to act as a pilot study in order to gather some preliminary data to: 1) assess whether the stimuli generated by *Pyllusion* behaves as expected for each of the 10 illusion types (i.e., increase of task difficulty and illusion strength leads to an increased error rate) and 2) develop a gauged understanding of the magnitude of effects, to then refine the stimuli parameters to a more sensible range (i.e., not overly easy and not impossibly hard).

@@

Fifty-two participants were recruited via *Prolific* ([www.prolificacademic.co.uk](www.prolificacademic.co.uk)), a popular crowd-sourcing platform to recruit subjects for online behavioural research. In a study comparing across different research recruitment platforms (e.g., MTurks, Qualtrics etc.), participants from *Prolific* were found to provide the highest data quality across all examined measures, including that of attention (the extent to which participants actually read the questions prior to responding), comprehension (the extent to which participants were able to understand the task instructions or items), honesty (the degree to which participants responded to questions truthfully) and reliability (the extent to which participants' responses are internally consistent within the same scale) [@peer2022]. For the purposes of our study, no restrictions on age or country were imposed on participant recruitment; the only requisite was a "fluent" proficiency in English to ensure that the task instructions, given in English, would be well-comprehended. Participants were incentivised with a reward of about \textsterling 7.5 for completing the task, which took an average of an hour to finish.

@@

Using the same recruitment procedure as in study 1, we recruited 256 participants, within which 6 were identified as outliers and excluded, leaving a final sample of 250 participants (Mean age = 26.5, SD = 7.6, range: [18, 69]; Sex: 48% females, 52% males). Please see study 3 for the full demographic breakdown. We discarded blocks with more than 50% of errors (2.16% of trials) and 0.76% trials with extreme response times (\< 125 ms or \> 4 SD above mean).

@@

The first part of the analysis focuses on modelling the effect of illusion strength and task difficulty on errors and reaction time, within each illusion. In order to achieve that, we will start by fitting General Additive Models (GAMs), which can accommodate possible non-linear effects and interactions. Errors were analyzed using Bayesian binomial mixed models, and reaction times (RT) of correct responses were analyzed using an ex-Gaussian family with the same fixed effects entered for the location $\mu$ (mean), scale $\sigma$ (spread) and tail-dominance $\tau$ of the RT distribution [@balota2011moving; @matzke2009psychological].

@@

<!-- This is likely to be caused by the outline. It will be fixed in the next release of the Pyllusion software. -->

Future studies should investigate reaction times to find the most appropriate models, and / or use models that integrate errors and reaction time (e.g., drift diffusion models).

@@

This study was based on the data collected in study 2. However, the variables of interest here were taken from the questionnaires that were inserted in between the two series of illusion blocks. We used the *IPIP6* (24 items; [@sibley2011]) to measure 6 personality traits (namely, extraversion, openness, conscientiousness, agreeableness, neuroticism and honesty-humility). On the other hand, the PID-5 (25 items; [@hopwood2012]) was used to measure the degree to which participants inhibit psychopathological personality traits (disinhibition, antagonism, detachment, negative affect and psychoticism). The participants were the same as in study 2. However, due to a technical issue, no data on the personality questions are available for the first eight participants.

@@

Using the parametric illusion generation framework we developed, Pyllusion [@makowski2021a], we have hence shown that illusions can be manipulated continuously across several different visual illusions. This opens the door for new illusions-based paradigms and tasks, therefore making it possible for future researchers to directly manipulate specific features and parameters of the illusion that are of interest. The validation of this novel framework also affords future illusion scientists a standardized measure of illusion susceptibility, instead of relying on conventional methods that depend upon participants' subjective perceptions. In our paradigm, in which we apply this approach to a reaction-time task, we were able to measure inter-individual scores of objective illusion sensitivity.

@@

We strongly invite researchers to explore and re-analyze our dataset with other approaches and methods to push the understanding of visual illusions and illusion sensitivity further. The task, data and analysis script are available in open-access at [**https://github.com/RealityBending/IllusionGameValidation**](https://github.com/RealityBending/IllusionGameValidation){.uri}.
